---
title: "Lab 7.2"
format: pdf
editor: visual
---

```{r}
install.packages("Lahman")
library(Lahman)
library(dplyr)
library(ggplot2)
data(Teams)
```

### Question 1:

```{r}
teams_recent <- Teams |>
  filter(yearID >= 2000, yearID != 2020)

dim(teams_recent)
```

-   The dimensions of the filtered dataset are 720 rows and 48 columns
-   The year 2020 should be excluded because the MLB season was severely shortened due to the COVID-19 pandemic. Instead of the standard 162-game season, teams only played 60 games in 2020. This makes the win totals from 2020 incomparable to other years

### Question 2:

```{r}
ggplot(teams_recent, aes(x = W)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Wins",
       x = "Wins", y = "Frequency")
```

-   The distribution is approximately normal (bell-shaped) and roughly symmetric, centered around 81 wins. Most teams win between 70-90 games, with very few teams at the extremes (fewer than 50 wins or more than 110 wins).
-   This matches expectations because baseball is a zero-sum game - every win requires another team to lose, forcing the league average to exactly .500 (81 wins out of 162 games). The bell shape reflects competitive balance in MLB, where most teams perform near average with progressively fewer extreme outcomes on either end.

### Question 3:

```{r}
ggplot(teams_recent, aes(x = R, y = W)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship Between Runs and Wins",
       x = "Runs Scored", y = "Wins")
```

Form: The relationship is linear.

Direction: The trend is positive.

Strength: The association appears to be strong. The data points cluster tightly around the blue trend line, indicating that runs scored are a good predictor of wins.

Outliers: There are few to no extreme outliers.

\- The initial prediction was correct. The relationship between W and R is linear, positive, and moderately dispersed. Teams scoring 500–600 points typically win about 50–70 games, while teams scoring over 900 points typically win over 100 games. The trend line in the scatter plot confirms this.

### Question 4:

```{r}
ggplot(teams_recent, aes(x = RA, y = W)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship Between Runs Allowed and Wins",
       x = "Runs Allowed", y = "Wins")
```

-   Form: Linear
-   Direction: Negative (more runs allowed, fewer wins)
-   Strength: Strong negative association
-   Outliers: A few present, particularly teams around 500-550 runs allowed with unusually high wins (110-115)
-   The relationships are mirror opposites. Both show similarly strong linear associations but in opposite directions: runs scored have a positive slope while runs allowed have a negative slope. The strength of association appears comparable between the two, suggesting that defense (preventing runs) is roughly as important as offense (scoring runs) in determining wins.

### Question 5:

```{r}
model_1 <- lm(W ~ R, data = teams_recent)
summary(model_1)
```

Equation: W=62.52  +  0.1538R  −  0.1007RA  −  0.00004075R²  +  0.0099HR

Interpretation: For each additional run scored in a season, we predict a team will win approximately (slope) more games on average

### Question 6:

```{r}
library(Lahman)
data(Teams)
teams_recent <- Teams |> filter(yearID >= 2000)
mean_runs <- mean(teams_recent$R, na.rm = TRUE)
mean_wins <- mean(teams_recent$W, na.rm = TRUE)
mean_runs
mean_wins

```

-   Average number of season runs is 719,2 and wins are 79

```{r}
model_1 <- lm(W ~ R, data = teams_recent)
predict(model_1, newdata = data.frame(R = mean_runs))
```

-   A team scoring the average number of runs is predicted to win approximately 79 games

### Question 7:

```{r}
predict(model_1, newdata = data.frame(R = 600))
predict(model_1, newdata = data.frame(R = 850))
predict(model_1, newdata = data.frame(R = 10000))
```

-   A team that scored 600 runs is predicted to win about 67 games, while a team with 850 runs is expected to win around 92 games-both are reasonable because they fall within the typical range of MLB team performance. However, a team with 10,000 runs is predicted to win over 1,000 games, which is clearly inaccurate since no team scores anywhere near that many runs, and a season only has 162 games. This occurs because the model is linear and continues to increase without bounds, resulting in unrealistic predictions when applied far outside the range of the training data. Linear models are reliable only within the scope of observed values.

### Question 8:

```{r}
model_2 <- lm(W ~ R + RA, data = teams_recent)
summary(model_2)
```

-   R² = 0.79

    Model_2 is substantially better than model_1:

-   Higher R²: 0.8875 vs model_1's R²

-   model_2 explains approximately 88.75% of the variance in wins

-   This makes sense because wins depend on both scoring runs (offense) AND preventing runs (defense)

-   Both coefficients are highly significant (p \< 2e-16)

-   Adding runs allowed (RA) captures the defensive aspect that was missing in model_1

### Question 9:

```{r}
model_3 <- lm(W ~ R + RA + I(R^2) + HR, data = teams_recent)
summary(model_3)
```

-   R² = 0.90
-   This model explains approximately 90% of the variance in team wins

### Question 10:

No, a positive coefficient does not prove causation, and sports management cannot use this model to draw causal conclusions. This is observational data, not from a randomized experiment, so there may be confounding variables (like player talent,team quality) that affect both runs scored and wins. Additionally, teams cannot manipulate runs scored independently of other factors - increasing runs requires changes in player performance, strategy, and roster decisions that are all interconnected. The model shows correlation and is useful for prediction, but cannot tell us what would happen from direct intervention on one variable.

### Last Question

Will you ensure that your submission to Gradescope…

is of a pdf generated from a qmd file, YES

has all of your code visible to readers YES

and assigns each of the questions to all pages that show your work for that question? YES
